---
title: "STAT 37810 Final Project #1: Metropolis-Hastings"
author: "Yingzhao Li & Linlin Wu"
date: "Nov. 1st, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1 Algorithm and implementation

```{r}
betametropolis <- function(startvalue, c, iterations=10000){
  chain = rep(0, iterations)
  for (i in 1:iterations){
    currentvalue <- rbeta(1, c * startvalue, c * (1 - startvalue))
    proposal <- dbeta(startvalue, c * currentvalue, c * (1 - currentvalue)) / dbeta(currentvalue, c * startvalue, c * (1 - startvalue))
    posterior <- dbeta(currentvalue, 6, 4) / dbeta(startvalue, 6, 4)
    if (runif(1) < min(1, posterior * proposal)){
      startvalue <- currentvalue
      chain[i] <- startvalue
    }else{
      chain[i] <- startvalue
    }
  }
  return(chain)
}

# generate a starting value from a uniform distribution
startvalue <- runif(1)
# set c=1 and run for 10,000 total iterations
draws <- betametropolis(startvalue, 1, 10000)
```

##2 Evaluate the performance of the sampler

### Trace plot, autocorrelation plot, and histogram of the draws

```{r}
# plots for c=1
par(mfrow = c(1, 3))
plot(draws, type = "l", main = "Trace plot"); acf(draws, main = "Autocorrelation plot"); hist(draws, main = "Histogram")
```

### Compare the draws with the target distribution Beta(6,4)

```{r}
#histogram of the draws with Beta(6,4)
par(mfrow = c(1, 1))
hist(draws, freq = FALSE)
x<-seq(0, 1, 0.01)
lines(x, dbeta(x, 6, 4))
```

We can see that graphically, the target distribution Beta(6,4) is not a very good fit for the histogram of the draws. 

```{r}
#ks test
ks.test(jitter(draws, 0.0001), rbeta(10000, 6, 4))
```

Most of the time, the p-value of the Kolmogorov–Smirnov test is less than 0.05. So, it's very likely that we will reject the null hypothesis that the draws and the target distribution Beta(6,4) are the same.

##3 Re-run this sampler with c = 0.1, c = 2.5 and c = 10

### Trace plot, autocorrelation plot, and histograms with c = 0.1, c = 2.5 and c = 10

```{r}
#plots for c = 0.1, 2.5, 10
cvec<-c(0.1, 2.5, 10)
for (i in 1:length(cvec)){
  par(mfrow=c(1, 3))
  draws=betametropolis(startvalue, cvec[i], 10000)
  plot(draws, type = "l", main = paste("Trace plot: c = ", cvec[i])); acf(draws, main = paste("Autocorrelation plot: c = ", cvec[i])); hist(draws, freq = FALSE, main = paste ("Histogram: c = ", cvec[i]))
  x<-seq(0, 1, 0.01)
  lines(x, dbeta(x, 6, 4))
}
```


From the autocorrelation plots, we can see that c = 1 has most lags, which requires thinning. c = 2.5 has least lags, which is most effective at drawing from the target distribution.

From the histograms, we can see that c = 10 seems to have the best fit for the target distribution Beta(6,4).

### The number of draws needed from the sampler before thinning and burn-in, and the comparison to the properties of the target distribution.
